---
title: "Soil carbon - fire frequency"
output:
  word_document:
    fig_height: 8
    fig_width: 8
  html_document: default
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

library(mxbutils)

library(dplyr)
library(ggplot2)

DAT.totalc <- loadFrom(mdPath("data/firefreq_totalc.RData"))
DAT.rpc <- loadFrom(mdPath("data/firefreq_rpc.RData"))

## ggplot bits

t <- theme_bw() +
  theme(text = element_text(size = 24))

theme_set(t)

PointSize <- 3

```


## Initial carbon

Distribution of percent initial carbon values for the depth and fire frequency categories
for which recalcitrant carbon has been assayed.

Horizontal lines indicate mean percent carbon value for each depth.


```{r}

dat <- DAT.totalc %>%
  filter(firefreq %in% c("1", "2", "4")) %>%
  filter(depth %in% c("0-5cm", "6-13cm"))

meanPercentC <- tapply(dat$percentC, dat$depth, mean, na.rm=TRUE)

ggplot(data = dat) + 
  geom_violin(aes(firefreq, percentC), 
              fill = "grey90") + 
  
  geom_point(aes(firefreq, percentC, colour = microsite), 
             size = PointSize, 
             position = position_jitter(width = 0.15, height = 0)) + 
  
  geom_abline(aes(intercept = meanPercentC[depth]), 
              slope = 0, colour = "brown", size = 1.5) + 
  
  labs(x = "Fire frequency", y = "%CTot") +
  
  facet_grid(depth ~ regionCode)


```


## Recalcitrant carbon value

Distribution of recalcitrant carbon values, calculated as mean of replicate digest values.

Horizontal lines indicate mean value for whole dataset.


```{r}

dat <- DAT.rpc %>%
  group_by(regionCode, depth, microsite, firefreq, replicate) %>% 
  summarize(digestPercentC = mean(digestPercentC, na.rm=TRUE))

meanPercentC <- tapply(dat$digestPercentC, dat$depth, mean, na.rm=TRUE)

ggplot(data = dat) + 
  geom_violin(aes(firefreq, digestPercentC), 
              fill="grey90") + 
  
  geom_point(aes(firefreq, digestPercentC, colour = microsite), 
             size = PointSize, 
             position = position_jitter(width=0.15, height=0)) +
  
  labs(x = "Fire frequency", y = "%RPC") +
  
  facet_grid(depth ~ regionCode)

```


## Recalcitrant carbon proportion

Distribution of recalcitrant carbon proportion values, calculated as mean of replicate digest values over initial carbon value.

Horizontal lines indicate mean proportion for whole dataset.


```{r}

dat <- DAT.rpc %>%
  group_by(regionCode, depth, microsite, firefreq, replicate) %>% 
  
  summarize(digestPercentC = mean(digestPercentC, na.rm=TRUE), 
            predigestPercentC = first(predigestPercentC)) %>%
  
  mutate(recalProp = digestPercentC / predigestPercentC)


ggplot(data = dat) + 
  geom_violin(aes(firefreq, recalProp), 
              fill = "grey90") +
  
  geom_point(aes(firefreq, recalProp, colour = microsite), 
             size = PointSize, 
             position = position_jitter(width=0.15, height=0)) +
  
  labs(x = "Fire frequency", y = "Prop RPC") +
  
  facet_grid(depth ~ regionCode)


```


